---
title: "Data_cleaning_thesis"
author: "Ilse Hamers"
date: "2026-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# readme
This script cleans and reshapes the data extracted directly from the lab system. Sample IDs are removed from this script to protect patient privacy.

## Libraries
```{r}
library(tidyr)
library(readr)
library(dplyr)
library(writexl)
library(readxl)
library(strinr)
```

## loading the data
```{r}
file_path <- Sys.getenv("ANEMIA_DATA_PATH", unset = "Data/anemie_data_extract_31DCE1D2.csv")

columns_to_keep <- c("Geslacht", "Leeftijd", "UniekLabnummer", 
                     "Testcode", "test", "resultaat", "eenheid", "Patientnummer")

data <- read_csv(file_path, col_select = all_of(columns_to_keep))
length(unique(data$UniekLabnummer))
```

## checking
```{r}
unique(data$test)
```

## filter relevant tests
```{r}
values_to_keep <- c(
  "HB",
  "LEUC",
  "CRPSE",
  "BSE", # bezinking
  "MCV",
  "ANE", # anemie protocol
  "NFERSE", # ferritine
  "VB12SE", # vitamine b12
  "HT", # hematocriet
  "ALTSE", # ALAT
  "KRESE", # kreatinine
  "NFOLSE", # foliumzuur
  "CKDESE", # eGFR
  "YZESE", # ijzer
  "TFSSE", # transf. sat.
  "TRFSE", # transferrine
  "TRC", # trombocyten
  "RTC", # reticulocyten relatief
  "PBNPSE", # NT-proBNP
  "LDSE", # LD
  "ARTC", # reticulocyten absoluut
  "ERY", # erytrocyten
  "HPTSE", # haptoglobine
  "OKC", # Opm.vakspec.
  "HOVS", #Opm. Laboratoriumspecialist
  "NTSHSE" # tsh
)

data_filtered <- data %>%
  filter(Testcode %in% values_to_keep) %>%
  select(-Testcode)

length(values_to_keep)
length(unique(data_filtered$test))
```

# Cases with anemia

```{r}
geen_anemie_ids <- data_filtered %>%
  filter(test == "Anemie protocol", resultaat == "Geen anemie.") %>%
  pull(UniekLabnummer) %>%
  unique()

data_anemie <- data_filtered %>%
  filter(!UniekLabnummer %in% geen_anemie_ids)

length(unique(data_anemie$UniekLabnummer))
```

## remove cases without usable results
```{r}
# cases without results
ongewenste_ids <- data_filtered %>%
  filter(test == "Anemie protocol", resultaat %in% c(
    "Geen materiaal ontvangen op het laboratorium.",
    "Materiaal gestold, aanvraag opnieuw insturen",
    "Materiaal gestold. Telefonisch gemeld",
    "Niet te beoordelen"
  )) %>%
  pull(UniekLabnummer) %>%
  unique()

data_filtered <- data_filtered %>%
  filter(!UniekLabnummer %in% ongewenste_ids)

length(unique(data_filtered$UniekLabnummer))

# cases where hb is not measured
geen_numeriek_hb_ids <- data_filtered %>%
  filter(test == "Hemoglobine") %>%
  filter(is.na(suppressWarnings(as.numeric(resultaat)))) %>%
  pull(UniekLabnummer) %>%
  unique()

data_filtered <- data_filtered %>%
  filter(!UniekLabnummer %in% geen_numeriek_hb_ids)

length(unique(data_filtered$UniekLabnummer))
```

```{r}
head(data_filtered)
```

# inspect all potential conclusion fields
```{r}
data_filtered %>%
  filter(test == "Anemie protocol") %>%
  count(resultaat, sort = TRUE)

data_filtered %>%
  filter(test == "Opm.vakspec.") %>%
  count(resultaat, sort = TRUE)

data_filtered %>%
  filter(test == "Opm. laboratoriumspecialist") %>%
  count(resultaat, sort = TRUE)
```

```{r}
# remove Anemie protocol = TEKST en (C)
data_filtered <- data_filtered %>%
  filter(!(test == "Anemie protocol" & resultaat %in% c("TEKST", "(C)")))

data_filtered %>%
  filter(test == "Anemie protocol") %>%
  count(resultaat, sort = TRUE)

# same for opm.vakspec.
data_filtered <- data_filtered %>%
  filter(!(test == "Opm.vakspec." & resultaat %in% c("TEKST", "(C)", "\n\r\r\n\r\r\n\r\r\n")))

data_filtered %>%
  filter(test == "Opm.vakspec.") %>%
  count(resultaat, sort = TRUE)

# ...and opm. laboratoriumspecialist
data_filtered <- data_filtered %>%
  filter(!(test == "Opm. laboratoriumspecialist" & resultaat %in% c("TEKST", "(C)")))

data_filtered %>%
  filter(test == "Opm. laboratoriumspecialist") %>%
  count(resultaat, sort = TRUE)
```

# check
```{r}
data_filtered %>%
  group_by(UniekLabnummer, test) %>%
  summarise(n_rijen = n(), .groups = "drop") %>%
  filter(n_rijen > 1) %>%
  arrange(UniekLabnummer, test)
```

# check if there's cases with both opm.vakspec. and opm. laboratoriumspecialist
```{r}
aantal_beide <- data_filtered %>%
  group_by(UniekLabnummer) %>%
  summarise(
    beide_aanwezig = all(c("Opm.vakspec.", "Opm. laboratoriumspecialist") %in% test)
  ) %>%
  filter(beide_aanwezig) %>%
  nrow()

aantal_beide
```

# merge double anemia protocol rows
```{r}
anemie <- data_filtered %>%
  filter(test == "Anemie protocol") %>%
  group_by(UniekLabnummer) %>%
  summarise(
    Geslacht   = first(Geslacht),
    Leeftijd   = first(Leeftijd),
    test       = first(test),
    resultaat  = paste(resultaat, collapse = " - "),
    .groups    = "drop"
  )

niet_anemie <- data_filtered %>%
  filter(test != "Anemie protocol")

data_filtered <- bind_rows(niet_anemie, anemie)
```

# check
```{r}
data_filtered %>%
  group_by(UniekLabnummer, test) %>%
  summarise(n_rijen = n(), .groups = "drop") %>%
  filter(n_rijen > 1) %>%
  arrange(UniekLabnummer, test)

data_filtered %>%
  group_by(UniekLabnummer, test) %>%
  summarise(n_rijen = n(), .groups = "drop") %>%
  filter(n_rijen > 1) %>%
  pull(test) %>% 
  unique()
```

#  fix up duplicates of other tests
```{r}
is_numeric_resultaat <- function(x) grepl("^\\s*-?\\d*[.,]?\\d+\\s*$", x)
bevat_cijfer <- function(x) grepl("\\d", x)

data_filtered <- data_filtered %>%
  group_by(UniekLabnummer, test) %>%
  mutate(
    heeft_cijfer = any(is_numeric_resultaat(resultaat)),
    heeft_tekst  = any(!is_numeric_resultaat(resultaat)),
    alle_tekst_zonder_cijfer = all(!bevat_cijfer(resultaat))
  ) %>%
  filter(
    # if both numerical and text: remove text row
    !(heeft_cijfer & heeft_tekst & !is_numeric_resultaat(resultaat))
  ) %>%
  mutate( # if exact duplicate rows: keep first row
    alle_identiek = n_distinct(resultaat) == 1
  ) %>%
  filter(!(alle_identiek & duplicated(resultaat))) %>%
  # if all rows are text: make first row = "Geen resultaat" and remove others
  mutate(
    resultaat = if_else(
      alle_tekst_zonder_cijfer & n() > 1 & row_number() == 1,
      "Geen resultaat",
      resultaat
    )
  ) %>%
  filter(!(alle_tekst_zonder_cijfer & n() > 1 & row_number() > 1)) %>%
  select(-heeft_cijfer, -heeft_tekst, -alle_tekst_zonder_cijfer, -alle_identiek) %>%
  ungroup()
```

# check
```{r}
data_filtered %>%
  group_by(UniekLabnummer, test) %>%
  summarise(n_rijen = n(), .groups = "drop") %>%
  filter(n_rijen > 1) %>%
  arrange(UniekLabnummer, test)
```

# trombocytes duplicates
(can have both a >x and a text field)
```{r}
data_filtered <- data_filtered %>%
  group_by(UniekLabnummer, test) %>%
  mutate(
    resultaat = if_else(
      test == "Trombocyten" & n() > 1,
      paste(
        resultaat[grepl("^>", resultaat)],
        resultaat[!grepl("^>", resultaat)],
        sep = " - "
      ),
      resultaat
    )
  ) %>%
  filter(!(test == "Trombocyten" & duplicated(UniekLabnummer))) %>%
  ungroup()
```

# Make one uniform conclusion column (Beschrijving)
(Opm.vakspec. else Anemie protocol --> or opm. laboratoriumspecialist else anemie protocol)
```{r}
beschrijving_rows <- data_filtered %>%
  filter(test %in% c("Anemie protocol", "Opm.vakspec.", "Opm. laboratoriumspecialist")) %>%
  group_by(UniekLabnummer) %>%
  summarise(
    resultaat = case_when(
      any(test == "Opm.vakspec." & !is.na(resultaat) & resultaat != "") ~ 
        resultaat[test == "Opm.vakspec." & !is.na(resultaat) & resultaat != ""][1],
      
      any(test == "Opm. laboratoriumspecialist" & !is.na(resultaat) & resultaat != "") ~ 
        resultaat[test == "Opm. laboratoriumspecialist" & !is.na(resultaat) & resultaat != ""][1],
      
      any(test == "Anemie protocol" & !is.na(resultaat) & resultaat != "") ~ 
        resultaat[test == "Anemie protocol" & !is.na(resultaat) & resultaat != ""][1],
      
      TRUE ~ NA_character_
    ),
    Geslacht = Geslacht[test == "Anemie protocol"][1],
    Leeftijd = Leeftijd[test == "Anemie protocol"][1],
    .groups = "drop"
  ) %>%
  mutate(test = "Beschrijving")

missing_cols <- setdiff(names(data_filtered), names(beschrijving_rows))
for (col in missing_cols) {
  beschrijving_rows[[col]] <- NA
}

beschrijving_rows <- beschrijving_rows[, names(data_filtered)]
data_filtered <- bind_rows(data_filtered, beschrijving_rows)
```
```{r}
# drop opmerkingen
data_filtered <- data_filtered %>%
 filter(!(test %in% c("Opm. laboratoriumspecialist", "Opm.vakspec.")))
```

```{r}
beschrijving_counts <- data_filtered %>%
  filter(test == "Beschrijving") %>%
  count(resultaat, sort = TRUE) %>%
  filter(n > 0)

beschrijving_counts
```

# Remove conclusion author names
Author names were seen in the following configurations:
- "Datum opmerking : ... Laboratoriumspecialist : ... ----------"
- Dr. of dr. of zonder punt
- J. Jansen of J van Jansen
- J. Jansen, AIOS Klinische Chemie
- J. Jansen Klinisch Chemicus
- J. Jansen, KCIO aan begin
- J. Jansen, AIOS

```{r}
# copy so can compare
data_zAuteur <- data_filtered
```

### 'Datum opmerking : .... Laboratoriumspecialist : ... ----------'
```{r}
patroon_Datum <- "(?s)^\\s*(Datum opmerking :|Datum consult :).*?\\-{10,}\\s*"

data_zAuteur$resultaat <- ifelse(
  data_zAuteur$test == "Beschrijving",
  str_replace(data_zAuteur$resultaat, patroon_Datum, ""),
  data_zAuteur$resultaat
)
```

### Eindigend met "Jan van Jansen, laboratoriumspecialist"
```{r}
patroon_specialist <- "\n[^\n]*[A-Z][a-z]+,\\s*laboratoriumspecialist(?s).*$"

data_zAuteur$resultaat <- ifelse(
  data_zAuteur$test == "Beschrijving",
  str_remove(data_zAuteur$resultaat, patroon_specialist),
  data_zAuteur$resultaat
)
```

### Dr. dr. en Dr 
```{r}
patroon_dr_punt <- "(?i)(\\n|\\.)\\s*dr\\.?\\s+[^\n]*[\\s\\r\\n]*$"

data_zAuteur$resultaat <- ifelse(
  data_zAuteur$test == "Beschrijving",
  str_remove(data_zAuteur$resultaat, patroon_dr_punt),
  data_zAuteur$resultaat
)
```

### klinisch chemicus en klinische chemie
```{r}
patroon_chemi_nl <- "(?i)\n.*(chemie|chemicus).*"

data_zAuteur$resultaat <- ifelse(
  data_zAuteur$test == "Beschrijving",
  str_remove(data_zAuteur$resultaat, patroon_chemi_nl),
  data_zAuteur$resultaat
)

patroon_chemi_direct <- "(?i)\\.[^.]*(chemie|chemicus)[\\s\\S]*$"

data_zAuteur$resultaat <- ifelse(
  data_zAuteur$test == "Beschrijving" & str_detect(data_zAuteur$resultaat, patroon_chemi_direct),
  str_replace(data_zAuteur$resultaat, patroon_chemi_direct, "."),
  data_zAuteur$resultaat
)
```

### , AIOS (eind) of , KCIO (start)
```{r}
patroon_aios_nl <- "(?i)\n.*AIOS.*"

data_zAuteur$resultaat <- ifelse(
  data_zAuteur$test == "Beschrijving",
  str_remove(data_zAuteur$resultaat, patroon_aios_nl),
  data_zAuteur$resultaat
)

patroon_kcio_start <- "(?i)^[^\n]*KCIO[^\n]*\n+"

data_zAuteur$resultaat <- ifelse(
  data_zAuteur$test == "Beschrijving",
  str_remove(data_zAuteur$resultaat, patroon_kcio_start),
  data_zAuteur$resultaat
)
```

# conclusions < 15 words removed
```{r}
uitzonderingen <- c("Geen anemie.")

ids_te_kort <- data_zAuteur %>%
  filter(test == "Beschrijving") %>%
  filter(!(resultaat %in% uitzonderingen) & str_count(resultaat, "\\S+") < 15) %>%
  pull(UniekLabnummer)

data_zAuteur <- data_zAuteur %>%
  filter(!UniekLabnummer %in% ids_te_kort)
```
```{r}
beschrijving_counts <- data_zAuteur %>%
  filter(test == "Beschrijving") %>%
  count(resultaat, sort = TRUE) %>%
  filter(n > 0)

beschrijving_counts
```

# to wide format
Unit added to test name so we can pivot
```{r}
data_zAuteur <- data_zAuteur %>%
  mutate(test = case_when(
    !is.na(eenheid) & eenheid != "" ~ paste0(test, " (", eenheid, ")"),
    TRUE ~ test
  ))

data_zAuteur <- data_zAuteur %>%
  select(-eenheid)
```

```{r}
data_wide <- data_zAuteur %>%
  pivot_wider(
    names_from = test, 
    values_from = resultaat
  )

data_wide<- data_wide %>%
  group_by(UniekLabnummer) %>%
  summarise(
    # take a not empty value to keep
    Leeftijd = first(na.omit(Leeftijd)),
    Geslacht = first(na.omit(Geslacht)),
    
    # not empty
    `Anemie protocol` = first(na.omit(`Anemie protocol`)),
    Beschrijving = first(na.omit(Beschrijving)),
    
    across(everything(), ~ first(na.omit(.x)))
  ) %>%
  ungroup()
```

## remove patient duplicates
```{r}
dubbele_patienten <- data_wide %>%
  count(Patientnummer) %>%
  filter(n > 1) %>%
  nrow()

dubbele_patienten

dubbele_patienten_geenanemie <- data_wide %>%
  group_by(Patientnummer) %>%
  filter(n() > 1) %>%
  filter(Beschrijving == "Geen anemie.") %>%
  ungroup() %>%
  distinct(Patientnummer) %>%
  nrow()

dubbele_patienten_geenanemie

data_wide <- data_wide %>%
  mutate(is_anemie = ifelse(Beschrijving == "Geen anemie.", 1, 0)) %>%
  group_by(Patientnummer) %>%
  arrange(is_anemie, desc(UniekLabnummer)) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  select(-is_anemie)

data_wide <- data_wide %>%
  select(-Patientnummer)
```



# export to csv
```{r}
write.csv(data_wide, "data_clean_wide.csv", row.names = FALSE)
```


# QLoRA training set export without test cases
```{r}
labnummers_lijst <- c(
) # censored

data_zTest <- data_wide %>%
  filter(!UniekLabnummer %in% labnummers_lijst) %>%
  select(-'Anemie protocol')

anemie_groep <- data_zTest %>% filter(Beschrijving != "Geen anemie.")
geen_anemie_groep <- data_zTest %>% filter(Beschrijving == "Geen anemie.")

set.seed(42)
geen_anemie_groep <- geen_anemie_groep %>% sample_n(53) # 10% van de dataset

data_QLoRA <- bind_rows(anemie_groep, geen_anemie_groep)
data_QLoRA <- data_QLoRA %>%
  mutate(Beschrijving = str_replace_all(Beschrijving, "\r\n|\r", "\n"))

```
```{r}
write_xlsx(data_QLoRA, "data_QLoRA.xlsx")
```







