---
title: "RAG_prep_thesis"
author: "Ilse Hamers"
date: "2026-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# readme
This script prepares the historic, gold standard, and test data for RAG/KNN.
- lab values are turned into numeric values
- values are converted into a direction class relative to reference range

## Libraries
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(readxl)
```

# Data
```{r}
data_dir <- Sys.getenv("ANEMIA_DATA_DIR", unset = "data")
path <- file.path(data_dir, "data_clean_wide.csv")

data <- read.csv(path, check.names = FALSE) %>%
  select(-`Anemie protocol`)

path_testcases <- file.path(data_dir, "10_testcases_wide.csv")
data_test <- read.csv(path_testcases, check.names = FALSE) %>%
  select(-Patientnummer, -`Anemie protocol`)
```

# exclude test cases from ref data
```{r}
test_ids <- c(
  
) # censored

data <- data %>%
  filter(!UniekLabnummer %in% test_ids)

write.csv(data, "data_originieel_z10_RAG.csv", row.names = FALSE)
```

# Age from 44j to 44
```{r}
data <- data %>%
  mutate(Leeftijd = as.numeric(str_remove(Leeftijd, "j")))

data_test <- data_test %>%
  mutate(Leeftijd = as.numeric(str_remove(Leeftijd, "j")))
```

# handle < and >
< x becomes half of x, > x becomes double of x
```{r}
parse_lab_value <- function(x) {
  case_when(
    str_detect(x, "^<") ~ as.numeric(str_remove(x, "^<")) / 2,
    str_detect(x, "^>") ~ as.numeric(str_remove(x, "^>")) * 2,
    TRUE ~ suppressWarnings(as.numeric(x))
  )
}

data <- data %>%
  mutate(across(
    .cols = where(is.character) & -all_of(c("Geslacht", "Beschrijving")),
    .fns  = parse_lab_value
  ))

data_test <- data_test %>%
  mutate(across(
    .cols = where(is.character) & -all_of(c("Geslacht", "Beschrijving")),
    .fns  = parse_lab_value
  ))

# strip unit from column name to match reference value table
clean_names <- function(x) {
  str_trim(str_remove(x, "\\s*\\(.*\\)"))
}
colnames(data)      <- clean_names(colnames(data))
colnames(data_test) <- clean_names(colnames(data_test))
```

# Load reference table
```{r}
path_ref <- file.path(data_dir, "Referentiewaarden.xlsx")
ref <- read_excel(path_ref)
```

# ref value range per sex into column
- x-x
- <x
- >x
```{r}
# upper = NA for "< x", lower = NA for "> x".
parseer_referentie <- function(ref_str) {
  ref_str <- str_trim(as.character(ref_str))
  if (is.na(ref_str) || ref_str == "" || ref_str == "NA") {
    return(c(NA_real_, NA_real_))
  }
  if (str_detect(ref_str, "^[0-9.,]+\\s*-\\s*[0-9.,]+$")) {
    delen <- str_split(ref_str, "\\s*-\\s*")[[1]]
    return(c(as.numeric(str_replace(delen[1], ",", ".")),
             as.numeric(str_replace(delen[2], ",", "."))))
  }
  if (str_detect(ref_str, "^<\\s*[0-9.,]+$")) {
    return(c(NA_real_,
             as.numeric(str_replace(str_remove(ref_str, "^<\\s*"), ",", "."))))
  }
  if (str_detect(ref_str, "^>\\s*[0-9.,]+$")) {
    return(c(as.numeric(str_replace(str_remove(ref_str, "^>\\s*"), ",", ".")),
             NA_real_))
  }
  c(NA_real_, NA_real_)
}

bouw_ref_tabel <- function(ref) {
  ref_tabel <- list()
  for (i in seq_len(nrow(ref))) {
    ref_tabel[[ref$Waarde[i]]] <- list(
      M = parseer_referentie(ref$Man[i]),
      V = parseer_referentie(ref$Vrouw[i])
    )
  }
  ref_tabel
}

ref_tabel <- bouw_ref_tabel(ref)
```

# Direction class: -2 / -1 / 0 / +1 / +2
```{r}
# thresholds "very low/high": gap > 30% (low) or > 50% (high) to reference.
# were chosen empirically with clinical chemist
THRESH_LOW  <- 0.30
THRESH_HIGH <- 0.50

direction_class <- function(v, lower, upper) {
  if (is.na(v)) return(NA_integer_)

  if (is.na(lower)) {                         # "< upper" eenzijdig
    if (v <= upper) return(0L)
    gap <- (v - upper) / upper
    return(if (gap > THRESH_HIGH) 2L else 1L)
  }
  if (is.na(upper)) {                         # "> lower" eenzijdig
    if (v >= lower) return(0L)
    gap <- (lower - v) / lower
    return(if (gap > THRESH_LOW) -2L else -1L)
  }
  if (v >= lower & v <= upper) return(0L)     # tweezijdig
  if (v < lower) {
    gap <- (lower - v) / max(lower, 1)
    return(if (gap > THRESH_LOW) -2L else -1L)
  } else {
    gap <- (v - upper) / max(upper, 1)
    return(if (gap > THRESH_HIGH) 2L else 1L)
  }
}
```

# Classify dataset
also keeps raw value in case of tie-rbeaking during KNN
```{r}
classificeer_dataset <- function(data, ref_tabel) {

  meta_kolommen <- c("UniekLabnummer", "Leeftijd", "Geslacht",
                     "Anemie protocol", "Beschrijving", "Source")
  lab_kolommen  <- setdiff(names(data), meta_kolommen)

  bekend <- intersect(lab_kolommen, names(ref_tabel))

  class_cols <- map_dfc(bekend, function(lab) {
    klassen <- map2_int(
      data[[lab]],
      data$Geslacht,
      function(v, s) {
        sleutel <- if (!is.na(s) && s == "M") "M" else "V"
        r       <- ref_tabel[[lab]][[sleutel]]
        direction_class(v, r[1], r[2])
      }
    )
    tibble(!!paste0(lab, "_klasse") := klassen)
  })

  raw_cols <- data %>% select(all_of(bekend))
  names(raw_cols) <- paste0(names(raw_cols), "_raw")

  bind_cols(
    data %>% select(UniekLabnummer, Geslacht),
    class_cols,
    raw_cols
  )
}
```

# apply to reference (historic) and test
```{r}
referentie_klassen <- classificeer_dataset(data, ref_tabel)
testcases_klassen <- classificeer_dataset(data_test, ref_tabel)

write.csv(referentie_klassen, "referentie_database_klassen_RAG.csv", row.names = FALSE)
write.csv(testcases_klassen, "10_testcases_klassen_RAG.csv", row.names = FALSE)
```

# check proportions
```{r}
klasse_kolommen <- names(referentie_klassen)[grepl("_klasse$", names(referentie_klassen))]

sanity <- map_dfr(klasse_kolommen, function(k) {
  x <- referentie_klassen[[k]]
  tibble(
    kolom = k,
    n = sum(!is.na(x)),
    pct_na = round(mean(is.na(x)) * 100, 1),
    pct_neg2 = round(mean(x == -2, na.rm = TRUE) * 100, 1),
    pct_neg1 = round(mean(x == -1, na.rm = TRUE) * 100, 1),
    pct_0 = round(mean(x == 0, na.rm = TRUE) * 100, 1),
    pct_pos1 = round(mean(x == 1, na.rm = TRUE) * 100, 1),
    pct_pos2 = round(mean(x == 2, na.rm = TRUE) * 100, 1)
  )
})

print(sanity)
```

# Gold standard
same thing
```{r}
path_goud <- file.path(data_dir, "Gouden_standaard.xlsx")
data_goud <- read_excel(path_goud) %>%
  select(-Patientnummer, -`Anemie protocol`)

colnames(data_goud) <- clean_names(colnames(data_goud))

# age to numeric
data_goud <- data_goud %>%
  mutate(Leeftijd = as.numeric(str_remove(.data$Leeftijd, "j")))

# < and >
data_goud <- data_goud %>%
  mutate(across(
    .cols = where(is.character) & -any_of(c("Geslacht", "Beschrijving", "UniekLabnummer")),
    .fns  = parse_lab_value
  ))

goud_klassen <- classificeer_dataset(data_goud, ref_tabel)

write.csv(goud_klassen, "data_goud_klassen_RAG.csv", row.names = FALSE)
```
