---
title: "data_exploration_thesis"
author: "Ilse Hamers"
date: "2026-04-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#readme
Data exploration of historic dataset

## Libraries
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(forcats)
library(scales)
library(purrr)
```

## Data
```{r}
data_dir <- Sys.getenv("ANEMIA_DATA_DIR", unset = "data")

path <- file.path(data_dir,"data_clean_wide.csv")
data <- read.csv(path, check.names = FALSE)

data <- data %>%
  mutate(Leeftijd = as.numeric(str_remove(Leeftijd, "j")))
```

---

# aantal consults
```{r, fig.width= 10, fig.height=4}
aanvragen <- read_excel(path = "C:\\Users\\ilseh\\Desktop\\Master thesis UHasselt\\Data\\aantallen.xlsx", col_names = T)

ggplot(aanvragen, aes(x = Startdatum, y = Aantal)) +
  geom_line(aes(colour = "#bc4242"), size = 1.5) +
  labs(title = "Number of anemia conclusions written at MUMC+ over time", x = "Date", y = "Amount") +
  theme_minimal() +
  theme(legend.position = "none")
```

# Sex distribution
```{r}
data %>%
  count(Geslacht) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

ggplot(data, aes(x = Geslacht)) +
  geom_bar(fill = "#bc4242") +
  labs(title = "Distribution of Sex (full dataset)", x = "Sex", y = "Count") +
  theme_minimal()
```

# Age distribution
```{r}
summary(data$Leeftijd)

ggplot(data, aes(x = Leeftijd)) +
  geom_histogram(binwidth = 5, fill = "#bc4242", color = "white") +
  labs(title = "Age Distribution (full dataset)", x = "Age", y = "Count") +
  theme_minimal()

ggplot(data, aes(x = Geslacht, y = Leeftijd, fill = Geslacht)) +
  geom_boxplot(alpha = 1) +
  scale_fill_manual(values = c("M" = "#bc4242", "V" = "#e8a0a0")) +
  labs(title = "Age by Sex (full dataset)", x = "Sex", y = "Age") +
  theme_minimal() +
  theme(legend.position = "none")
```

# Anemia percentage
```{r}
data <- data %>%
  mutate(anemie = !is.na(`Anemie protocol`) & `Anemie protocol` != "Geen anemie.")

data %>%
  count(anemie) %>%
  mutate(
    label = ifelse(anemie, "Anemia", "No anemia"),
    pct   = round(n / sum(n) * 100, 1)
  )
```

# same plots but for anemia subset
```{r}
data_anemie <- data %>%
  filter(`Anemie protocol` != "Geen anemie.")

# Sex
data_anemie %>%
  count(Geslacht) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

ggplot(data_anemie, aes(x = Geslacht)) +
  geom_bar(fill = "#bc4242") +
  theme_minimal() +
  labs(title = "Distribution of Sex (anemia cases only)", x = "Sex", y = "Count")

# Age
summary(data_anemie$Leeftijd)

ggplot(data_anemie, aes(x = Leeftijd)) +
  geom_histogram(binwidth = 5, fill = "#bc4242", color = "white") +
  theme_minimal() +
  labs(title = "Age Distribution (anemia cases only)", x = "Age", y = "Count")

ggplot(data_anemie, aes(x = Geslacht, y = Leeftijd, fill = Geslacht)) +
  geom_boxplot(alpha = 1) +
  scale_fill_manual(values = c("M" = "#bc4242", "V" = "#e8a0a0")) +
  labs(title = "Age by Sex (anemia cases only)", x = "Sex", y = "Age") +
  theme_minimal() +
  theme(legend.position = "none")
```

# Conclusion (Beschrijving) word count
```{r}
data_anemie <- data_anemie %>%
  mutate(beschrijving_words = str_count(Beschrijving, "\\S+"))

summary(data_anemie$beschrijving_words)

ggplot(data_anemie, aes(x = beschrijving_words)) +
  geom_histogram(binwidth = 5, fill = "#bc4242", color = "white") +
  theme_minimal() +
  labs(
    title = "Word Count of Expert Conclusions (Beschrijving)",
    x = "Number of Words",
    y = "Count"
  )
```

# Missingness per lab variable
```{r}
meta_cols <- c("UniekLabnummer", "Geslacht", "Leeftijd",
               "Anemie protocol", "Beschrijving", "anemie", "Patientnummer")
lab_cols  <- setdiff(names(data), meta_cols)

missingness <- data %>%
  select(all_of(lab_cols)) %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  arrange(desc(pct_missing))

print(missingness, n = Inf)

ggplot(missingness, aes(x = reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "#bc4242") +
  coord_flip() +
  labs(
    title = "Missingness per Lab Variable (full dataset)",
    x = "Variable",
    y = "% Missing"
  ) +
  theme_minimal()
```

```{r fig.height=7}
missingness_split <- data %>%
  select(anemie, all_of(lab_cols)) %>%
  group_by(anemie) %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100), .groups = "drop") %>%
  pivot_longer(-anemie, names_to = "variable", values_to = "pct_missing") %>%
  mutate(anemie = ifelse(anemie, "Anemia", "No anemia"))

ggplot(missingness_split,
       aes(x = reorder(variable, pct_missing), y = pct_missing, fill = anemie)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Anemia" = "#bc4242", "No anemia" = "#e8a0a0")) +
  labs(
    title = "Missingness per Lab Variable by Anemia Status",
    x = "Variable",
    y = "% Missing",
    fill  = NULL
  ) +
  theme_minimal()

missingness_split %>%
  pivot_wider(names_from = anemie, values_from = pct_missing) %>%
  arrange(desc(Anemia)) %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))
```

# Lab value distributions (numeric variables) --> not used in thesis

## by sex
```{r fig.height=4, fig.width=6, results='asis'}
walk(unique(data_long$variable), function(v) {
  p <- data_long %>%
    filter(variable == v) %>%
    ggplot(aes(x = value, fill = Geslacht)) +
    geom_density(alpha = 0.7) +
    scale_fill_manual(values = c("M" = "#bc4242", "V" = "#e8a0a0")) +
    labs(title = v, x = "Value", y = "Density", fill = "Sex") +
    theme_minimal()
  print(p)
})
```

## by anemia status
```{r fig.height=4, fig.width=5, results='asis'}
walk(unique(data_long$variable), function(v) {
  p <- data_long %>%
    filter(variable == v) %>%
    ggplot(aes(x = anemie, y = value, fill = anemie)) +
    geom_boxplot(outlier.size = 0.5, alpha = 0.8) +
    scale_x_discrete(labels = c("FALSE" = "No anemia", "TRUE" = "Anemia")) +
    scale_fill_manual(values = c("FALSE" = "#e8a0a0", "TRUE" = "#bc4242")) +
    labs(title = v, x = NULL, y = "Value") +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)
})
```

## Direction-class distribution (from RAG prep output)
```{r fig.height=7, fig.width=9}
path_klassen <- "C:\\Users\\ilseh\\Desktop\\Master thesis UHasselt\\Scripts\\referentie_database_klassen_RAG.csv"
ref_klassen  <- read.csv(path_klassen, check.names = FALSE)

klasse_cols <- names(ref_klassen)[grepl("_klasse$", names(ref_klassen))]

klasse_long <- ref_klassen %>%
  select(all_of(klasse_cols)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "klasse") %>%
  mutate(
    variable = str_remove(variable, "_klasse$"),
    klasse   = factor(klasse, levels = c(2, 1, 0, -1, -2))
  ) %>%
  filter(!is.na(klasse))

klasse_pct <- klasse_long %>%
  count(variable, klasse) %>%
  group_by(variable) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

var_order <- klasse_pct %>%
  filter(klasse == 0) %>%
  select(variable, pct_normal = pct)

var_order_full <- klasse_pct %>%
  distinct(variable) %>%
  left_join(var_order, by = "variable") %>%
  mutate(pct_normal = replace_na(pct_normal, 0))

klasse_pct <- klasse_pct %>%
  left_join(var_order_full, by = "variable")

ggplot(klasse_pct,
       aes(x = reorder(variable, pct_normal), y = pct, fill = klasse)) +
  geom_col(position = "stack") +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "2" = "#7b0000",
      "1" = "#e06060",
      "0" = "#d0d0d0",
      "-1" = "#6080c0",
      "-2" = "#1a3a7a"
    ),
    labels = c("+2 Very high", "+1 High", "0 Normal", "-1 Low", "-2 Very low")
  ) +
  labs(
    title = "Direction-Class Distribution per Lab Variable",
    x = "Variable",
    y = "% of non-missing values",
    fill  = "Direction class"
  ) +
  theme_minimal()
```


